---
title: "Get Started with ESM Package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get_Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Loading the needed packages

```{r setup}
library(ESM)
library(terra)
```

## Case 1: Modeling with presence and absence data

### Data preparation

Here we first need to load our data. To realize an ESM, we first gather some information: the coordinates of the observations (called xy); a response vector (resp) attributing 1 if the species was present or 0 if absent; the name of the species (sp.name) and; environmental predictors (env), which can be a SpatRaster from terra package or a data.frame.

Let's focus here to model the distribution of *Tayloria serrata*, which were only found 15 times out of 413 plots, in the Western Swiss Alps. Here, we will use 5 predictors, obtained from the ecospat package and projected into the same CRS as our observations.

We first need to download and prepare our data :

```{r}
## Species information with environmental values
data("ESM_species.env") 

## SpatRaster with 5 environmental predictors
data("ESM_Env")

xy <- ESM_species.env[,1:2] #coords
resp <- ESM_species.env[,3] #occurences
sp.name <- "Tayloria_serrata" #species selected
env  <- terra::unwrap(ESM_Env) #the environment 

```

Let's take a look at our data while plotting the observations across the study area

```{r echo=TRUE, warning=FALSE}
terra::plot(terra::subset(env, 1), col = "grey85", 
            legend = FALSE, box = FALSE, axes = FALSE)
terra::points(xy[resp ==0,], pch = 13, col = "firebrick")
terra::points(xy[resp ==1,], pch = 16, col = "black")
```

### Modeling Core

The first step to model species distribution is to select the modelling technique you will used. With the package ESM, you can modify the parameters of each of the 5 modelling techniques available, with the function *ESM_Models.Options*. For this example, we will perform GLM with a binomial family, allowing linear and quadratic terms and performing a step AIC to select the best structure.

```{r echo=TRUE}
mod.options <- ESM_Models.Options(GLM = list(type = "quadratic",
                                             test = 'AIC'))
```

Once the modelling technique is selected, the next step is to generate bivariate models using the function *ESM_Modeling*, allowing by default all the possible pairs of predictors.

```{r}
my.ESM <- ESM_Modeling(resp = resp,
                       xy = xy,
                       env = env,
                       sp.name = sp.name,
                       models = c("GLM"),
                       models.options = mod.options,
                       prevalence = 0.5,
                       cv.method = "split-sampling",
                       cv.rep = 10,
                       cv.ratio = 0.7,
                       parallel = FALSE,
                       save.models = FALSE,
                       save.obj = FALSE,
                       verbose = FALSE)
```

Once the models are generated, *ESM_Modeling* also calculates the predictions and the evaluation of the calibration and the transferability of all the bivariate models :

```{r}
# Prediction for one bivariate model
head(my.ESM$biva.predictions$ddeg0.mind68)

my.ESM$biva.calibration #is the model able to correctly predict the probability of occurrence inside the environment in which it was calibrated (70% of our data if ratio = 0.7).

my.ESM$biva.evaluations#is the model able to correctly predict the probability of occurrence of new points based on new environmental data (30% of our data if ratio = 0.7).
```

Finally, you can use the function *ESM_Projection* to predict new probability of occurrence of *Tayloria serrata* in a completely new environment. The function will produce raster maps (.tif) that will be stocked in a new folder of your computer.

```{r}
proj <- ESM_Projection(ESM.Mod = my.ESM, 
                       new.env = env,
                      name.env = "current", 
                      parallel = FALSE, 
                      save.obj = FALSE)

Proj_TS <- rast(list.files(paste0(getwd(),"/ESM.output_Tayloria_serrata/current"), 
                           full.names=T)) #load all the maps from the new folder

plot(Proj_TS[[1]]) #plot the first climatic map
```

Now, in practice, you will not use these bivariate models to make predictions. The objective of the ESMs is to generate bivariate models to overcome the overfitting problem related to the low number of occurrences and then combine them. To do so, you will have to use the function *ESM_Ensemble.Modeling* that will generates an ensemble model combining all the bivariate ones based on their respective performances.

In this case, the max TSS was chosen to represent the performance of each model but you're free the chose any other metrics.

```{r}
my.ESM_EF <- ESM_Ensemble.Modeling(my.ESM, 
                                   weighting.score=c("MaxTSS"), 
                                   threshold=0,
                                   save.obj = FALSE)
```

Then, as above, the function *ESM_Ensemble.Projection* will allow you to predict new probability of occurrence for our species of interest based on the ensemble model which combine all the variables in our environment !

```{r}
Ens.proj <- ESM_Ensemble.Projection(ESM.proj = proj,
                                    ESM.ensembleMod = my.ESM_EF,
                                    save.obj = FALSE)

plot(Ens.proj)
```

Secondly, you can determine a threshold that will set the limit between your presences and absences among the probabilities of occurrence using the function *my.ESM_thresholds*. Once the threshold is determine, you can project a binary map based on this one using *my.ESM_EFproj_current_binary*. In this case, the one based on the max TSS of calibrated ESMs into a new space was used.

```{r}
my.ESM_thresholds <- ESM_Threshold(my.ESM_EF)

my.ESM_EFproj_current_binary <- ESM_Binarize(Ens.proj,
                                             1000*my.ESM_thresholds$TSS.th)
plot(my.ESM_EFproj_current_binary)
```

```{r}
## get the variable contributions of ESMs

ESM_Variable.Contributions(my.ESM,
                           my.ESM_EF)

## get the response plots of ESMs

my.ESM_responsePlot <- ESM_Response.Plot(my.ESM,
                                         my.ESM_EF, 
                                         fixed.var.metric = 'mean') 
```